Sketch Recognition using Particle Swarm Algorithms

### 2009-07-26 trunk/smallpaper/SmallPaper.tex
making the paper 4 pages

\documentclass{article}
\usepackage{spconf}
\usepackage{amsmath}
\usepackage[dvips,pdftex]{graphicx}
\usepackage{subfigure}
\usepackage{named}
\usepackage[dvips,pdftex]{hyperref}
\title{Sketch Recognition using Particle Swarm Algorithms}
%\sthanks{Professor}\sthanks{Associate Professor}\sthanks{Professor}\sthanks{Associate Professor}
\twoauthors
  {Maha El Meseery, Mahmoud Fakhr El Din, Samia Mashali}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, samia@eri.sci.eg}
  {Magda Fayek, Nevin Darwish}
	{ Computer Engineering Department\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.  
\end{abstract}
\section{Introduction}
The field of symbol recognition has gained interest in the last few years. Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of a few advanced sketch recognition systems.

Sketch recognition is divided into three main steps preprocessing, segmentation and symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke then proceeds to remove noise and compute basic statistical and geometrical information. In segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier system.

@@

This paper introduces a new sketch recognition system, which uses particle swarm algorithm (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. Segmentation is based on dividing the digital curve into polygon\cite{PolygonApproximationPSO}. The algorithm enhances its performance by adding curvature and other local information while segmenting strokes to achieve better results. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a symbol features vector composed of different statistical, geometrical and spatial features is deduced. The final classification step, SVM classifier uses the computed feature vector to classify the segments into one of the previously trained classes.

@@

\section{The proposed system}
\label{Sysdisc}
 The suggested sketch recognitions system is divided into three main steps 1) preprocessing, 2) segmentation and 3) Recognition. The following section provides a detailed description of each step.
\subsection{Preprocessing}
\label{Prepross}
 Speed and time difference information were widely used in sketch understanding systems \cite{earlyprocess}. It was noted that time difference provides more distinct maxima than the speed information. Agar \cite{polygonfeedback31} mentions that the pointing device (for example the mouse) sampling rate is the reason for this phenomena. 
 In the presented system, we computed time difference, direction, speed and curvature of each point along a stroke. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ is calculated as the angle between two vectors and curvature is considered as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.  Dominant points are characterized by lower speed values and higher curvature and direction values.

After the system computes all the speed, time difference and curvature information it proceeds to detect the points with low velocity and high curvatures. Using simple differentiation to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted a process presented by \cite{earlyprocess}, where the mean of the curve is calculated. Then a threshold \textit{th} is used to separate the curve into regions $Region_i$; each region $Region_i$ is defined as a range of points, where the curve values are either above or below the threshold \textit{th}. Those regions are further processed to find the maximum point $Max(Region_i)$ of each region $Region_i$. The stroke points $p_i(x,y)$ that correspond to those maximum values are labeled as \textit{possible dominant points} $P_{pd}$. The system repeats this process for curvature, time difference and speed information. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. Figure \ref{fig:LabelsPPD} shows the particles labeled as Possible dominate point $P_{pd}$ by the preprocessing, it is noticed the redundancy of some $P_{pd}$ points. % (as shown are redundant)
\begin{figure}
	\centering
		\subfigure[Posible domninate point]{ 	\label{fig:LabelsPPD}	\includegraphics[scale=0.5]{images/ppd.jpg}}
		\hfill
			\subfigure[DPSO algorithm encoding]{	\label{fig:LabelsPSO} \includegraphics[scale=0.5]{images/pso1.jpg}}			
	\caption{Segmentation}% a) Possible dominate point b) Particle encoding  
	\label{fig:pso1}
\end{figure}
\subsection{Segmentation}
\label{seg}
In the segmentation stage the stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm \cite{chernov}. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two particle swarm algorithms that divide the stroke to either lines or lines and curves. The algorithms takes the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing then produce a set of dominant points which are connected by either lines or curves.  The next section describes the ellipse detection algorithm and the two particle swarm algorithms used to divide the stroke.
 \begin{figure}
	\centering
		\includegraphics[scale=0.55]{images/segblock.jpg}
	\caption{Segmentation Block}
	\label{fig:segblock}
\end{figure}
\subsubsection{Ellipse Detection} 
The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation \cite{chernov}.  
\subsubsection{Discrete particle swarm algorithm (\textit{DPSO})}
\label{subsubsec:Discreteparticleswarmalgorithm}
Two DPSO algorithms are used to generate segmentations for each stroke the user draws. For each generated segmentation evaluated an error is determined. The segmentation with the minimum error value is chosen as the best stroke segmentation (Figure \ref{fig:segblock}). The problem definition is the same in both algorithms but they differ in the method they use to compute fitness and error functions.

@@

For the \textit{first algorithm \textsl{AlgS1}}, 
 the arc $\widehat{x_lx_k}$ is defined as the consecutive set of points from point$x_l$ to point $x_{k}$ as in $x_l,x_{l+1} \cdots,x_k$. The line $\overline{x_l x_k} $ as the straight line connecting point $x_l$ to point $x_k$. The approximation error is computed by the equation \ref{eq:ErrorSwarm1}. The graphical meaning of the error is shown in Figure\ref{fig:LabelsPSO}.
\begin{equation}
E=\sum\nolimits_{l = 1}^M e ( \widehat{x_lx_{l+1}},\overline{x_l x_{l+1}})
\label{eq:ErrorSwarm1}
\end{equation} where $M$ is the number of dominant points in this solution as generated by the swarm algorithm. The error $e ( \widehat{x_lx_k},\overline{x_l x_k})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_lx_k}$ to the line $\overline{x_l x_k}$. The fitness is computed using equation \ref{eq:fitnessSwarm1}\begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   { - E/\varepsilon N} & {ifE > \varepsilon ,}  \\
   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\
\end{array}} \right.
\label{eq:fitnessSwarm1}
\end{equation} where $N$ is the number of points in the stroke, $D$ is the number of points in the solution that was previously labeled as a possible dominant point ($P_{pd}$), $E$ is the computed error and $\varepsilon$ is the error threshold. It should be noticed that when the error is larger than the threshold $\varepsilon$ the fitness is given a $-ve$ value to lower the fitness value of the solution. Otherwise the system favors the lower number of vertices.

@@

\section{Conclusion and Future Work}
\label{ConclusionandFutureWork}
This paper presented a new approach to sketch recognition using DPSO. The system uses multiple preliminary information, i.e., speed and curvature information, as input to the \textit{DPSO}. Results show that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The final recognition is based on geomterical,statistical and structural features. The experiments evaluated two \textit{DPSO} different segmentation algorithms (AlgS1 and AlgS2) on simple presentation set dataset. Results show that AlgS2 achieves better segmentation and final accuracy than Alg3 \cite{earlyprocess} on the dataset used. The results proved that the system is easily expandable to more and more symbols as it does not depend on low level primitive recognizers but on high level symbol and geometrical features.  The system achieved an average overall improvement of more than 2\% over Alg3.  
%This paper presented a new approach to sketch recognition using PSO. The system uses both speed and curvature information which help improving the \textit{DPSO}. It is noted that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The tradeoff between accuracy achieved and time complexity must be further investigated to achieve better results. The use of statistical moments and structural features improves the recognition rate. The system was tested on 13 different symbols and achieved an accuracy of 91.5\%. The system does not depend on low level recognizer but rather on a set of high level features. This makes the system easily expandable to more and more symbols.

 A possible extension of this research is to complete the clustering algorithm for fully automated sketch recognition. Currently, the system process one symbol at a time, an enchantment would be to allow users to draw more than one symbol at a time and incorporate a method for separating symbols. Other area of enhancements is the features extraction methods. Introducing more spatial and geometrical features is believed to improve classifications.  
\bibliographystyle{IEEEbib}
\bibliography{../../neededfiles/Bibliographies/Mybibliography}
\end{document}

>>>>

\documentclass{article}
\usepackage{spconf}
\usepackage{amsmath}
\usepackage[dvips,pdftex]{graphicx}
\usepackage{subfigure}
\usepackage{named}
\usepackage[dvips,pdftex]{hyperref}
\title{Sketch Recognition using Particle Swarm Algorithms}
%\sthanks{Professor}\sthanks{Associate Professor}\sthanks{Professor}\sthanks{Associate Professor}
\twoauthors
  {Maha El Meseery, Mahmoud Fakhr El Din, Samia Mashali}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, samia@eri.sci.eg}
  {Magda Fayek, Nevin Darwish}
	{ Computer Engineering Department\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.  
\end{abstract}
\section{Introduction}
The field of symbol recognition has gained interest in the last few years. Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of a few advanced sketch recognition systems.Sketch recognition is the process of identifying user strokes into a meanful symbol that can be furthur used.  
%Sketch recognition is divided into three main steps preprocessing, segmentation and symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke then proceeds to remove noise and compute basic statistical and geometrical information. In segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier system.

@@

This paper introduces a new sketch recognition system, which uses particle swarm algorithm (PSO) to optimally segment user's strokes. Users can draw symbols using any number of strokes in any order they like. The system segments each stroke using two different PSO algorithms. Segmentation is based on dividing the digital curve into polygon\cite{PolygonApproximationPSO}. To handle complex strokes, our system uses a second segmentation PSO algorithm to segment strokes into a set of lines and circular arcs. After segmentation, a symbol features vector composed of different statistical, geometrical and spatial features is deduced. The final classification step, SVM classifier uses the computed feature vector to classify the segments into one of the previously trained classes.

@@

\section{The proposed system}
\label{Sysdisc}
 The suggested sketch recognitions system is divided into three main steps 1) preprocessing, 2) segmentation and 3) Recognition. The following section provides a detailed description of each step.
\subsection{Preprocessing}
\label{Prepross}
 Speed and time difference information were widely used in sketch understanding systems \cite{earlyprocess}. It was noted that time difference provides more distinct maxima than the speed information. Agar \cite{polygonfeedback31} mentions that the pointing device (i.e. the mouse) sampling rate is the reason for this phenomena. 
 In the presented system, we computed time difference, direction, speed and curvature of each point along a stroke. The speed is calculated as $v=\Delta s/\Delta t$ where $t$ is the time difference between two points and $s$ is the length between them. The direction $d$ is calculated as the angle between two vectors and curvature is considered as the change in direction with respect to length i.e. $c= \Delta d/\Delta s$.  Dominant points are characterized by lower speed values and higher curvature and direction values.

After the system computes all the speed, time difference and curvature information it proceeds to detect the points with low velocity and high curvatures. Using simple differentiation to detect local extreme points resulted in false points due to the non smooth curves. Hence, the system adopted a process presented by \cite{earlyprocess}, where the mean of the curve is calculated. Then a threshold \textit{th} is used to separate the curve into regions $Region_i$; each region $Region_i$ is defined as a range of points, where the curve values are either above or below the threshold \textit{th}. Those regions are further processed to find the maximum point $Max(Region_i)$ of each region $Region_i$. The stroke points $p_i(x,y)$ that correspond to those maximum values are labeled as \textit{possible dominant points} $P_{pd}$. The system repeats this process for curvature, time difference and speed information. All the points labeled as possible dominant points $P_{pd}$ are saved into a single array. Figure \ref{fig:LabelsPPD} shows the particles labeled as Possible dominant point $P_{pd}$ by the preprocessing, it is noticed the redundancy of some $P_{pd}$ points. % (as shown are redundant)
\begin{figure}
	\centering
		\subfigure[Posible domninate point]{ 	\label{fig:LabelsPPD}	\includegraphics[scale=0.5]{images/ppd.jpg}}
		\hfill
			\subfigure[DPSO algorithm encoding]{	\label{fig:LabelsPSO} \includegraphics[scale=0.5]{images/pso1.jpg}}			
	\caption{Segmentation}% a) Possible dominate point b) Particle encoding  
	\label{fig:pso1}
\end{figure}
\subsection{Segmentation}
\label{seg}
In the segmentation stage the stroke is divided into a set of primitives. As shown in Figure \ref{fig:segblock} first an attempt is made to fit the stroke points into a curve or an ellipse using a minimum square error fitting algorithm. If the stroke proved to be an ellipse arc then the segmentation process ends and the system proceeds to the next step. Otherwise, the stroke is passed to two particle swarm algorithms that divide the stroke to either lines or lines and curves. The algorithms takes the stroke points along with the possible dominant points $P_{pd}$ computed during preprocessing then produce a set of dominant points which are connected by either lines or curves.  The next section describes the ellipse detection algorithm and the two particle swarm algorithms used to divide the stroke.
 \begin{figure}
	\centering
		\includegraphics[scale=0.55]{images/segblock.jpg}
	\caption{Segmentation Block}
	\label{fig:segblock}
\end{figure}
\subsubsection{Ellipse Detection} 
The process starts by computing the center of the stroke bounding box. The bounding box center point is used as the first estimation of the ellipse center. The axes of the ellipse are estimated as $width/2$ and $height/2$ of the stroke bounding box. The least square fitting algorithm is used to minimize the fitting error of the ellipse equation.  
\subsubsection{Discrete particle swarm algorithm (\textit{DPSO})}
\label{subsubsec:Discreteparticleswarmalgorithm}
Two DPSO algorithms are used to generate segmentations for each stroke the user draws. For each generated segmentation evaluated an error is determined. The segmentation with the minimum error value is chosen as the best stroke segmentation (Figure \ref{fig:segblock}). The problem definition is the same in both algorithms but they differ in the method they use to compute fitness and error functions.

@@

For the \textit{first algorithm \textsl{AlgS1}}, 
 the arc $\widehat{x_lx_k}$ is defined as the consecutive set of points from point$x_l$ to point $x_{k}$ as in $x_l,x_{l+1} \cdots,x_k$. The line $\overline{x_l x_k} $ as the straight line connecting point $x_l$ to point $x_k$. The approximation error is computed by the equation \ref{eq:ErrorSwarm1}. The graphical meaning of the error is shown in Figure\ref{fig:LabelsPSO}.
\small\begin{equation}
E=\sum\nolimits_{l = 1}^M e ( \widehat{x_lx_{l+1}},\overline{x_l x_{l+1}})
\label{eq:ErrorSwarm1}
\end{equation} 
\normalsize
where $M$ is the number of dominant points in this solution as generated by the swarm algorithm. The error $e ( \widehat{x_lx_k},\overline{x_l x_k})$ is computed as the sum of squared perpendicular distance from every point along the arc $\widehat{x_lx_k}$ to the line $\overline{x_l x_k}$. The fitness is computed using equation \ref{eq:fitnessSwarm1}
\small \begin{equation}
\max fitness(p_i ) = \left\{ {\begin{array}{*{20}c}
   { - E/\varepsilon N} & {ifE > \varepsilon ,}  \\
   {D/\sum\limits_{j = 1}^N {p_{ij} } } & {otherwise}  \\
\end{array}} \right.
\label{eq:fitnessSwarm1}
\end{equation}\normalsize where $N$ is the number of points in the stroke, $D$ is the number of points in the solution that was previously labeled as a possible dominant point ($P_{pd}$), $E$ is the computed error and $\varepsilon$ is the error threshold. It should be noticed that when the error is larger than the threshold $\varepsilon$ the fitness is given a $-ve$ value to lower the fitness value of the solution. Otherwise the system favors the lower number of vertices.

@@

\section{Conclusion and Future Work}
\label{ConclusionandFutureWork}
This paper presented a new approach to sketch recognition using DPSO. Results show that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The final recognition is based on geomterical,statistical and structural features. The experiments evaluated two \textit{DPSO} different segmentation algorithms (AlgS1 and AlgS2) on simple presentation set dataset. Results show that AlgS2 achieves better segmentation and final accuracy than Alg3 \cite{earlyprocess} on the dataset used. The results proved that the system is easily expandable to more and more symbols as it does not depend on low level primitive recognizers but on high level symbol and geometrical features.  The system achieved an average overall improvement of more than 2\% over Alg3.  
%This paper presented a new approach to sketch recognition using PSO. The system uses both speed and curvature information which help improving the \textit{DPSO}. It is noted that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The tradeoff between accuracy achieved and time complexity must be further investigated to achieve better results. The use of statistical moments and structural features improves the recognition rate. The system was tested on 13 different symbols and achieved an accuracy of 91.5\%. The system does not depend on low level recognizer but rather on a set of high level features. This makes the system easily expandable to more and more symbols.

 A possible extension of this research is to complete the clustering algorithm for fully automated sketch recognition. Currently, the system process one symbol at a time, an enchantment would be to allow users to draw more than one symbol at a time and incorporate a method for separating symbols. Other area of enhancements is the features extraction methods. Introducing more spatial and geometrical features is believed to improve classifications.

\bibliographystyle{IEEEbib}
\bibliography{../../neededfiles/Bibliographies/Mybibliography}
\end{document}
